---
title: "How can we deal with missing data in R?"
author: ""
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = FALSE)
```

## What are NAs?

Basic vocabulary:

* `NA` stands for *not available*
* `NaN` stands for *not a number*
* `Inf` stands for *infinite*
* `NULL` stands for a *null object* 

But really... what are NAs?

* NAs are **placeholders**

## Some R quirks

Most important thing to remember: NAs are **contagious**!

```{r quirk1}
NA > 1
NA/3
NA == NA # This last one is important!
```

## Some R quirks (2)

Which items are missing? Which are not?
```{r quirk2.1, results='hold'}
vec <- c("R", "ladies", NA, "Paris")

# Won't work
vec == NA 
vec != NA
```

```{r quirk2.2, results='hold'}
# Yey!
is.na(vec)
!is.na(vec)
```

## Some R quirks (3)

```{r quirk3, results='hold'}
vect <- c(2, 2, 2, NA)

sum(vect)
sum(vect, na.rm = TRUE)
```


## How can we deal with missing values?

*Ignore* the missing values and work only with complete cases

* Lose key information, bias your analysis
* Values may be missing for a reason!
  
*Impute* the missing values

* Lots of methods!
* But other shortcomings

## Before treatment... exploration

  * How many are there?
  * Where are the missing values?
  * Are they related?
  * Can I make assumptions to help with the imputation?
  
## Simple example with `airquality`

```{r libs, message=FALSE, warning=FALSE}
# Libraries

library(dplyr)
library(zoo)   # locf imputation
library(VIM)   # visualization

# Data
head(airquality)
```

## Some info on the data

```{r exploration}
summary(airquality)
```

## Exploration

```{r rship1}
airquality %>% 
  group_by(Month) %>% 
  summarise(
    miss_ozone = sum(is.na(Ozone)), 
    miss_solar = sum(is.na(Solar.R)),
    miss_both  = sum(is.na(Ozone) & is.na(Solar.R)), 
    n_month    = n()
    )
```

## Exploration (2)

```{r rship2}
airquality %>% 
  select(Ozone, Solar.R) %>% 
  marginplot()
```

## Imputation

```{r impute1}
# mean imputation with dplyr
airquality %>% 
  mutate_at(
    .cols = vars(Solar.R, Ozone),
    .funs = funs(ifelse(is.na(.), mean(., na.rm = T), .))
    ) %>% 
  head()
```

## Imputation (2)

```{r impute2}
# mean imputation with zoo
airquality %>% 
  na.aggregate() %>% 
  head()
```

## Imputation (3)

```{r impute3}
# locf imputation with zoo
airquality %>% 
  na.locf() %>% 
  head()
```


## Links & packages

More on visualization with VIM:
<https://cran.r-project.org/web/packages/VIMGUI/vignettes/VIM-Imputation.pdf)>

"Tagged" missing values (importing from STATA and SPSS):
<http://haven.tidyverse.org/reference/tagged_na.html>

Summary of different R packages for imputation
<https://www.rstudio.com/rviews/2016/11/30/missing-values-data-science-and-r/)>

More on imputation methods (in French & with some math): 
<http://www.math.univ-toulouse.fr/~besse/Wikistat/pdf/st-m-app-idm.pdf>